---
title: "sequential-trial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sequential-trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This vignette describes the use of the functions `data_splitter()` and `trial_constructor()`. These functions are the core elements of the package `seqtrial`. Their goal is to reformat a survival data into a 'counting process' format and to construct a sequence of trials from that. Typically, these functions are used is a subsequent manner.

For the construction of a sequence of trials from observational data, survival data needs to be in a 'counting process' format. A counting process form of data allows analyses of data that contain fixed variables and time-varying variables. Typical examples of time-varying variables in our setting are treatment status and confounding variables that influence the treatment decision and the outcome. An example of a fixed variable is sex. The function `data_splitter()` splits a survival data set into multiple subrecords at time intervals of length 1 for the duration of follow-up and adds a column indicating treatment status in each of the intervals.

In a sequential trial, a trial is constructed at each day of follow-up. Each of these trials consists of two arms: a treated arm and an untreated arm. In trial 0, all individuals who have started treatment on day 0 (day zero) are included in the treated arm; all individuals who have not started treatment on day 0 (but may start treatment later during follow-up) are included in the untreated arm. In trial 1, all individuals who have started treatment on day 1 are included in the treatment arm; all individuals who have not started treatment by day 1 (but may start treatment later during follow-up) are included in the untreated arm. Individuals who had started treatment on day 0 are excluded from trial 1 and later trials. The same logic applies for subsequent trials. The function `trial_constructor()` constructs trials from data in a counting process format.

In this vignette, you will learn the key ideas behind `data_splitter()` and `trial_constructor()`.

To begin, you will need to load the `seqtrial` and `dplyr` package.
```{r setup}
library(seqtrial)
library(dplyr)
```

# Split Data
`data_splitter()` splits survival data set by creating new rows of intervals of length 1 for the duration of follow-up. To accommodate the creation of these new rows, the new colums `tstart` and `tend` will be added. For follow-up equal to `fup`, a new row will be added for intervals 0-1, 1-2, ..., 2-`fup`. In the last row `fup`-1-`fup`, the `status` column will indicate if an outcome was experienced or not. The time-varying `treatment` column will indicate if treatment was started in a interval. It is assumed that there is only one switch, that is, once treatment is started, it is continued.

## Example
Data set `paxlovid` comprises two individuals followed-up for `fup` days. The `status` column indicates if an outcome was experienced at the end of follow-up. Both individuals have started treatment `A` after `tt_A` days (time-to-treatment). Column `L` indicates the fixed variable sex.
```{r paxlovid-example}
paxlovid
```

To split into a 'counting process' format, we use `data_splitter()`:
```{r paxlovid-splitter}
paxlovid %>%
  data_splitter(treatment = A,
                tt_treatment = tt_A)
```

- The first argument is the data to be split, `paxlovid`. 
- `treatment` describes the treatment column in `data`, in this case that is `A`. 
- `tt_treatment` describes the column denoting time-to-treatment, in this case that is `tt_A`.
- Note that the arguments `id`, `fup`, `status` were not used since the default values for these columns could be used (which are `id`, `fup` and `status`, all existent in `paxlovid`). 

## Note on time-to-treatment variable
Note that time-to-treatment indicates the number of days (or weeks/months/years if another scale is used) without treatment. In other words, 0 indicates that there were 0 intervals without treatment and treatment was started in the first interval, 1 indicates that the first interval was without treatment and treatment was started in interval two, 2 indicates the first two interval were without treatment, .., t indicates the first t intervals were without treatment and treatment was started in day t:t+1. 

In contrast, a fup of t days with status indicator 1 means that an outcome was experienced in interval t-1:t (assuming right-closed intervals).

# Construct Trials
`trial_constructor()` helps constructing trials once data has been split and is in 'counting process' format. The constructed trials start in the first interval and end at `n-1`-`n`.

## Example
```{r paxlovid-trialconstructor}
trials <- 
  paxlovid %>%
    data_splitter(treatment = A,
                  tt_treatment = tt_A) %>%
    trial_constructor(n = 5,
                      treatment = A)
trials
```

- The first argument is the data split using `data_splitter()`.
- `n` describes the number of trials to construct.
- `treatment` describes the treatment column in `data_splitted`.
- Note that the arguments `id` and `status` were not used since the default values for these values could be used (which are `id` and `status`, all existent in the splitted `paxlovid` data, see above).
- `censor` indicates if individuals in the no-treatment arm of an trial are censored when they start treatment.

The resulting tibble comprises 5 trials. Individual 1 starts treatment after 4 intervals and individual 2 starts treatment after 3 intervals. This means that in the first 3 trials, individuals 1 and 2 are both in the untreated arm. In trial 4, individual 1 is in the untreated arm and individual 2 in the treated arm. In the fifth trial, both individuals are in the treated arm.

```{r paxlovid-trials}
trials %>%
  group_by(trial, arm) %>%
  summarise(n = id %>% unique() %>% length(),
            .groups = "keep")
```

## Deviation from baseline treatment status
Since both individuals start treatment during follow-up, they deviate from their baseline treatment in some of the constructed trials. `trial_constructor()` can censor the individuals as soon as they deviate from their baseline treatment. For example, in trial 1, individual 1 starts treatment after 4 intervals and therefore only the first 4 intervals are included in the no-treatment arm. Individual 2 starts treatment after 3 intervals and therefore only the first 3 intervals are included in the no-treatment arm of trial 1.
```{r paxlovid-trialconstructor-censored}
trials_censored <- 
  paxlovid %>%
    data_splitter(treatment = A,
                  tt_treatment = tt_A) %>%
    trial_constructor(n = 5,
                      treatment = A,
                      censor = TRUE)
trials_censored %>%
  filter(trial == 1) %>%
  group_by(id, arm) %>%
  summarise(n_interval = n(),
            .groups = "keep")
```
In trial 4, individual 1 only contributes their observed interval 3-4 in the untreated arm and individual 2 contributes intervals 3-18 (which equals 15 intervals). Note that `tstart` and `tend` in trial 4 start again at 0 and 1, respectively: interval 0-1 in trial 4 is interval 3-4 in trial 1.
```{r paxlovid-trialconstructor-censored}
trials_censored %>%
  filter(trial == 4) %>%
  group_by(id, arm) %>%
  summarise(n_interval = n(),
            .groups = "keep")
```
